Synthesizing Quotient Lenses

### 2018-02-22 papers/QuotientLenses/main.tex
qres

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\kw}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2
\kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonizer}{\ensuremath{\kw{Canonizer}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}

@@

We propose to do away with much of this overhead and the burden of a
fiddly programming process by defining a single new compact notation,
the \emph{Quotient Regular Expression}, which simultaneously describes
external format, equivalence relation and canonical forms, and then to
synthesize lenses that map between these objects automatically.

@@

A {\em Quotient Regular Expression} (QRE) defines an external regular
expression $R$,
an equivalence relation $\sim$ on the language described by $R$, and a canonical set
of representatives of $\sim$, all in one syntactic term.
Figure~\ref{fig:example-qre} illustrates how to 
specify \bibtex{} records and their canonical elements using QREs.

@@

In Figure~\ref{fig:example-qre}, the QRE \verb+project [ \n\t\r]+$^{+} \to$
\verb+" "+ matches any non-zero number of whitespace characters and quotients
them to a single equivalence class with a single space as the canonical
representative. We call this QRE a \textit{projection} because an entire regular
language is projected to a single representative string.

Another QRE in Figure~\ref{fig:example-qre} is given by

\begin{center}
\begin{lstlisting}
squash comma_name $\to$ space_name using (get space_to_comma)
\end{lstlisting}
\end{center}

\noindent This QRE is a \textit{squash} QRE. A \textit{squash} QRE ``glues''
two data formats using a user-defined function. For example,
the QRE in the code above specifies that the \cd{comma_name} representation of
the author's name is equivalent to the \cd{space_name} representation, with
\cd{space_name} taken to be the canonical representation. The
function that identifies the two formats is given by \cd{get space_to_comma},
which transforms names in the \cd{comma_name} representation to names in the
\cd{space_name} representation. In this example, \cd{space_to_comma} is itself
an automatically \textit{synthesized} lens between the \cd{space_name} and
\cd{comma_name} QREs, and we need to use the ``get'' direction of the lens to
specify the squash transformation.

The third example of a QRE in the Figure is \cd{bib_fields}, defined by:
\begin{center}
\begin{lstlisting}
perm (label, bib_author, bib_title) with (project ("," . wsp$^+$ ) $\to$ ",\n")
\end{lstlisting}
\end{center}

\noindent This is a \textit{permutation} QRE that specifies that the label,
title, and author fields can occur in any order, with the fields separated by a
comma followed by arbitrary amounts of whitespace. The canonical representation
is that in which the label, author and title fields occur in that order,
separated by a comma followed by a newline. The \textit{permutation} QRE is
designed for equivalences such as this, where different permutations of
equivalences are interspersed with a separator which itself has an
equivalence defined on it.

Given the \textit{projection}, \textit{squash}, and \textit{permutation}
QRE primitives, we can build more complex languages using the usual regular
expression combinators of alternative choice `\cd{|}', concatenation `\cd{.}'
and Kleene star `$^*$', and also sequential composition ``;''.

>>>>

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\kw}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2
\kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonizer}{\ensuremath{\kw{Canonizer}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}

@@

We propose to do away with much of this overhead, and the burden of a
fiddly programming process, by defining a single new compact notation
to specify data formats modulo an equivalence relation and then by
synthesizing the lenses that map between these formats automatically.

@@

A {\em Quotient Regular Expression} (QRE) defines an external regular
expression $R$, an equivalence relation $\sim$ on the language
described by $R$, and a canonical set of representatives of $\sim$,
all at once.  QREs are constructed from a simple set of primitives
and composed using familiar regular operators
(choice `\cd{|}', concatenation `\cd{.}', sequential composition ``;''
and Kleene star `$^*$').

For example, to begin construction of our \bibtex{} format, we might
first define a whitespace format, which, externally, matches
any non-zero number of whitespace characters, and converts
any such whitespace into a single space character (its canonical form).
Such a definition uses the \kw{project} primitive to implement the
specification:

\begin{lstlisting}
let wsp = project [ \n\t\r]$^+$ $\to$ " "
\end{lstlisting}

While a projection chooses a single canonical representative for all elements
of a regular language, the \kw{squash} combinator maps data from one
regular language to another via a user-defined
function.  For instance,
if \codefont{comma_name} describes the format for ``Conway, J. H.''
and \codefont{space_name} describes the format for ``J. H. Conway''
and \codefont{s_to_c} is a function from one to the other then
the following instance of squash creates the desired canonizer.

%Figure~\ref{fig:example-qre} illustrates
%how to specify \bibtex{} records and their canonical elements using
%QREs.

\begin{lstlisting}
let name = squash comma_name $\to$ space_name using s_to_c
\end{lstlisting}

One way to define the function used by squash is simply to write it
from scratch in OCaml.  However, our system will also synthesize
such functions automatically from pairs of regular languages and
examples.
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Conway, J. H.", "J. H. Conway")}
let s_to_c = get l
\end{lstlisting}
%
The first line above synthesizes a \emph{bidirectional} lens between
\codefont{comma_name} and \codefont{space_name} using the listed example
transformation as a guide.  The second line simply
extracts the left-to-right transformation from the lens, which is what we need for squash.

A third useful QRE primitive is the permutation combinator.  For example,
the following instance of \kw{perm} allows label, author and title fields
(which we assume have been defined earlier) to appear in any order.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) 
\end{lstlisting}
%
To normalize any repeated separator between fields, one can extend a permutation
with any other QRE.  For instance, below, we normalize whitespace between fields,
leaving only a single newline.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) with (project ("," .wsp$^+$) $\to$ ",\n")
\end{lstlisting}
%
\noindent
The QRE for the bibtex format is presented in its entirety in Figure~\ref{fig:example-qre}.

@@

%% In Figure~\ref{fig:example-qre}, the QRE \verb+project [ \n\t\r]+$^{+} \to$
%% \verb+" "+ matches any non-zero number of whitespace characters and quotients
%% them to a single equivalence class with a single space as the canonical
%% representative. We call this QRE a \textit{projection} because an entire regular
%% language is projected to a single representative string.

%% Another QRE in Figure~\ref{fig:example-qre} is given by

%% \begin{center}
%% \begin{lstlisting}
%% squash comma_name $\to$ space_name using (get space_to_comma)
%% \end{lstlisting}
%% \end{center}

%% \noindent This QRE is a \textit{squash} QRE. A \textit{squash} QRE ``glues''
%% two data formats using a user-defined function. For example,
%% the QRE in the code above specifies that the \cd{comma_name} representation of
%% the author's name is equivalent to the \cd{space_name} representation, with
%% \cd{space_name} taken to be the canonical representation. The
%% function that identifies the two formats is given by \cd{get space_to_comma},
%% which transforms names in the \cd{comma_name} representation to names in the
%% \cd{space_name} representation. In this example, \cd{space_to_comma} is itself
%% an automatically \textit{synthesized} lens between the \cd{space_name} and
%% \cd{comma_name} QREs, and we need to use the ``get'' direction of the lens to
%% specify the squash transformation.

%% The third example of a QRE in the Figure is \cd{bib_fields}, defined by:
%% \begin{center}
%% \begin{lstlisting}
%% perm (label, bib_author, bib_title) with (project ("," . wsp$^+$ ) $\to$ ",\n")
%% \end{lstlisting}
%% \end{center}

%% \noindent This is a \textit{permutation} QRE that specifies that the label,
%% title, and author fields can occur in any order, with the fields separated by a
%% comma followed by arbitrary amounts of whitespace. The canonical representation
%% is that in which the label, author and title fields occur in that order,
%% separated by a comma followed by a newline. The \textit{permutation} QRE is
%% designed for equivalences such as this, where different permutations of
%% equivalences are interspersed with a separator which itself has an
%% equivalence defined on it.

%% Given the \textit{projection}, \textit{squash}, and \textit{permutation}
%% QRE primitives, we can build more complex languages using the usual regular
%% expression combinators of alternative choice `\cd{|}', concatenation `\cd{.}'
%% and Kleene star `$^*$', and also sequential composition ``;''.
